Controlled Experiments for Word Embeddings

### 2015-10-08 article/main.tex
related work to that of schnabel et al 2015

In this paper, we wish to study in detail to what extent these two
factors determine word vectors.  For a given corpus, both term
frequency and co-occurrence are, of course, fixed and it is not obvious
how to unravel these dependencies in an unambiguous, objective manner.
In particular, it is difficult to establish the distinctiveness of the
contexts in which a word is used.  To overcome these problems, we
propose to modify the training corpus in a controlled fashion.  To
this end, we insert new tokens into the corpus with varying frequencies
and varying levels of noise in their co-occurrence distributions.  By
modeling the frequency and co-occurrence distributions of these tokens
on existing words in the corpus, we are able to study their effect on
word vectors independently of one another.  We can thus study a family
of tokens that all appear in the same context, but with different
frequencies, or study a family of tokens that all have the same
frequency, but appear in a different number of contexts.  Starting from
the limited number of contexts in which a word appears in the original
corpus, we can increase this number by interspersing the word in arbitrary
contexts at random.  The word thus looses its significance in a
controlled way.  Although we present our approach using the word2vec
CBOW model, these and related experiments could equally well be carried
out for other word embedding methods such as the word2vec skip-gram
model \cite{DistRepns,EfficientEstimation}, GloVe
\cite{pennington2014glove}, and SENNA \cite{collobert-2011}.

We show that the length of the word vectors generated by the CBOW model
depends more or less linearly on both word frequency and level of noise
in the co-occurrence distribution of the word.  In both cases, the
coefficient of linearity depends upon the word.  If the co-occurrence
distribution is fixed, then word vector length increases with word
frequency.  If, on the other hand, word frequency is held constant, then
word vector length decreases as the level of noise in the co-occurrence
distribution of the word is increased.  We show furthermore that the
direction of the word vectors is independent of word frequency and only
moderately depends upon co-occurrence noise.  When noise is added to the
co-occurrence distribution of a word, the corresponding vector smoothly
interpolates between the original word vector and a small vector
perpendicular to it that represents a word with pure noise in its
co-occurrence distribution.  Surprisingly, this special vector in the feature
space, obtained by interspersing a token uniformly at random throughout
the corpus with a frequency sufficiently large to sample all contexts, is
non-zero.

@@

\begin{section}{Related work}\label{related-work}
\mar{move to other paper?}
Our notion of word significance is reminiscent of the concept of
``contextual distinctiveness'' introduced earlier by McDonald and
Shillcock \cite{mcdonald2001contextual}.  They define the contextual
distinctiveness of a target word $w$ as the Kullback-Leibler divergence
between the unconditional word distribution $\p(w')$, and the
co-occurrence distribution $\p(w' | w)$ of the target word $w$.  So
defined, contextual distinctiveness is a measure of the amount of
information provided by word $w$ about its contexts of use.  More
specifically, if a word $w$ tends to appear in a wide variety of
contexts, then its co-occurrence distribution tends to diverge less from
unconditional word distribution, and its contextual distinctiveness is
accordingly small.  If, on the other hand, $w$ typically appears in only
a small number of different contexts, its co-occurrence distribution
tends to diverge more from the unconditional word distribution, and a
larger value of contextual distinctiveness results.  To reduce the
computational burden, the authors considered only $690$ target words,
randomly chosen from equally divided frequency bands.  Furthermore, the
authors considered only co-occurrence with the $500$ most frequent
content words.  This significant restriction, necessary in their case in
order to reduce the variance of the estimates of their measure of
contextual distinctiveness, does not apply in our study.

@@

\begin{section}{Corpus and model}\label{corpus-and-model}
Our training data is built from the Wikipedia data dump from October
2013.  To remove the bulk of robot-generated pages from the training
data, only pages with at least 20 monthly page views are
retained.\footnote{For further justification and to obtain the dataset,
  see\\ \blogpost} Stubs and disambiguation pages are also removed,
leaving 463 thousand pages with a total of 482 million words.
Punctuation marks and numbers were removed from the pages and all words
were lower-cased.  Word frequencies are summarised in
Table~\ref{word-occurrences-table}.  This base corpus is then modified
as described in Sections~\ref{WFVE} and \ref{CNVE}.  For
recognisability, the tokens inserted into the corpus are upper-cased.

@@

\begin{subsection}{Replacement procedure}\label{replacement-procedure}
In the experiments detailed below, we modify the corpus in a controlled
manner by introducing tokens into the corpus via a replacement
procedure.  For the frequency experiment, the procedure is as follows.
Consider a word, say \word{cat}.  For each occurrence of this word, a
sample $i$, $1 \leqslant i \leqslant n$ is drawn from a truncated
geometric distribution, and that occurrence of the word \word{cat} is
replaced with the token \word{CAT\_i}.  In this way, the word \word{cat}
is replaced throughout the corpus by a family of tokens with varying
frequencies but approximately the same co-occurrence distribution as
\word{cat}.  That is, all these tokens are used in roughly the same
contexts as the original word.

The geometric distribution is truncated to limit the number of tokens
inserted into the corpus.  For any choice $0 < p < 1$ and maximum value
$n > 0$, the truncated geometric distribution is given by the
probability density function
\begin{equation} 
\label{distro}
P_{p, n} (i) = \frac{p^{i-1}  (1-p)}{1 - p^n}, \qquad 1
\leqslant i \leqslant n.
\end{equation} 
The factor in the denominator, which tends to unity in the limit $n \to
\infty$, assures proper normalisation.  We have chosen this distribution
because the probabilities decay exponentially base $p$ as a function of
$i$.  Of course, other distributions might equally well have been chosen
for the experiments.

@@

\begin{section}{Varying word frequency}\label{WFVE}
In this first experiment, we investigate the effect of word frequency on
the word embedding.  Using the replacement procedure, we introduce a
small number of families of tokens into the corpus.  The tokens in each
family vary in frequency but, replacing a single word, all share a
common co-occurrence distribution.  This allows us to study the role of
word frequency in isolation, everything else being kept equal.  We
consider two types of tokens.

\begin{subsection}{Tokens derived from existing words}\label{WFVEexisting}
We choose uniformly at random a small number of words from the
unmodified vocabulary for our experiment.  In order that the inserted
tokens do not have too low a frequency, only words which occur at least
10 thousand times are chosen.  We also include the high-frequency
stopword \word{the} for comparison.  Table~\ref{word-frequency-counts}
lists the words chosen for this experiment along with their frequencies.

The replacement procedure of Section~\ref{replacement-procedure} is then
performed for each of these words, using a geometric decay rate of $p =
\tfrac{1}{2}$, and maximum value $n=20$, so that the 1st token is
inserted with a probability of about $0.5$, the 2nd with a probability of
about $0.25$, and so on.  This value of $p$ is one of a range of values
that ensure that, for each word, multiple tokens will be inserted with a
frequency sufficient to survive the low-frequency cut-off of
128.  A maximum value $n=20$ suffices for this choice of $p$, since
$2^{20 + \log_2{128}}$ exceeds the maximum frequency of any token in
the corpus.  Figure~\ref{fig:word-frequency-experiment-text-cat}
illustrates the effect of these modifications on a sample text, with a
family of tokens \word{CAT\_i}, derived from the word \word{cat}.
Notice that all occurrences of the word \word{cat} have been replaced
with the tokens \word{CAT\_i}.

@@

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{Example sentences modified in the word frequency
          experiment as per Section~\ref{WFVEexisting}, where the word
          \word{cat} is replaced with tokens \word{CAT\_i} using the
          truncated geometric distribution (\ref{distro}) with
          $p=\tfrac{1}{2}$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}
\label{WFVEmeaningless}
Whereas the tokens introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we intersperse the word uniformly at random throughout the corpus
so that its relative frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the unconditional word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure~\ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher relative frequency
of $0.05$ is used instead for illustrative purposes.  Although the probability of inserting token
\word{VOID\_1} is twice that of inserting token \word{VOID\_2} and four
times that of inserting token \word{VOID\_3}, these three tokens all
appear once in this small piece of text.

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{The same example sentences as in
          Figure~\ref{fig:word-frequency-experiment-text-cat} where
          instead of the word \word{cat} now the meaningless word
          \word{VOID} is replaced with tokens \word{VOID\_i}.  For
          illustrative purposes, the meaningless word \word{VOID} was
          here interspersed with a relative frequency of
          $0.05$. }
	\label{fig:word-frequency-experiment-text-void}
\end{figure}
\end{subsection}

@@

\subsubsection{Word frequency and vector direction}\label{WFVE-direction}
Figure~\ref{word-frequency-experiment-heatmap} shows the cosine
similarity of pairs of vectors representing some of the tokens used in
this experiment.  Recall that the cosine similarity measures the extent
to which two vectors have the same direction, taking a maximum value of
$1$ and a minimum value of $-1$.  The number of different tokens
associated with an experiment word is the number of times that its
frequency can be halved and remains above the low-frequency cut-off of
$128$.

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and so a
limited number of samples is required to capture their co-occurrence
distributions.  We therefore expect that the vectors representing the
tokens derived from any of these words can be properly learned from the
modified corpus, even though their combined frequency equals the
frequency of that word in the original corpus.  The red blocks in
Figure~\ref{word-frequency-experiment-heatmap} clearly demonstrate that
the direction of these vectors is constant for each family of tokens.
In other words, the direction of these vectors is only determined by the
original word and is independent of how often that word appears,
provided sufficient samples have been processed for the word vector to
settle in its equilibrium direction.

The artificial word \word{VOID}, on the other hand, occurs by
construction in every context.  A much higher number of samples is
therefore required to capture its co-occurrence distribution, and
thereby to learn its word vector.  The same is true, but to a lesser
extent, for the stopword \word{the}.  The difficulty of learning the
vector representing \word{VOID\_i} or \word{THE\_i} increases with $i$,
for the tokens occur increasingly less often.  This is consistent with
Figure~\ref{word-frequency-experiment-heatmap}, where it is apparent
that the vectors representing \word{VOID\_i} for small $i$, say $i = 1,
\dots, 5$, roughly share a common direction, while this is no longer
true for higher values of $i$.  The same trend is apparent, although
less extremely so, for the word \word{the}.  The vectors for the tokens
\word{THE\_i} share a common direction for values of $i$ as high as
$14$, before they start changing direction.  We conclude that these
deviations are artifacts of poor learning due to an insufficient number
of occurrences of these tokens.
Table~\ref{word-frequency-experiment-cosine-similarity}, showing the
most similar words to each token \word{THE\_i}, supports this further.
In the discussions below, the tokens \word{VOID\_i} and \word{THE\_i}
whose vectors have a cosine similarity less than $0.8$ with
\word{VOID\_1} and \word{THE\_1}, respectively, are therefore discarded.

@@

\subsubsection{Word frequency and vector length}
We next consider the effect of frequency on word vector length.
Throughout, we measure vector length using the Euclidean norm.
Figure~\ref{fig:word-frequency-experiment-graph} shows this relation for
individual words, both for the word vectors, represented by the weights
of the first synaptic layer, syn0, in the word2vec neural network, and
for the vectors represented by the weights of the second synaptic layer,
syn1neg.  We include the latter, which are typically ignored, for
completeness.  Each line corresponds to a single word, and the points on
each line indicate the frequency and vector length of the tokens derived
from that word.  For example, the six points on the line corresponding
to the word \word{protestant} are labeled, from right to left, by the
tokens \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots,
\word{PROTESTANT\_6}.  Again, the number of points on the line is
determined by the frequency of the original word.  For example, the
frequency of the word \word{protestant} can be halved at most $6$ times
so that the frequency of the last token is still above the low-frequency
cut-off.  Because all the points on a line share the same co-occurrence
distribution, the left panel in
Figure~\ref{fig:word-frequency-experiment-graph} demonstrates
conclusively that length does indeed depend on frequency directly.
Moreover, this relation is seen to be approximately linear for each word
considered.  Notice also that the relative positions of the lengths of
the word vectors associated with the experiment words are roughly
independent of the frequency band, i.e., the plotted lines rarely cross.

Notice the inflection of the syn1neg curve in
Figure~\ref{fig:word-frequency-experiment-graph} for \word{the} at
approximately $10^6$.  We have no explanation of this.  Also observe
that the lengths of the vectors representing the meaningless tokens
\word{VOID\_i} are approximately constant (about $2.5$).  Since we
already found the direction to be also constant, it is sensible to speak
of the word vector of \word{VOID} irrespective of its frequency.  In
particular, the token \word{VOID\_1} may be taken as an approximation.

\begin{figure}
	\includegraphics[scale=0.5]{word-frequency-experiment-heatmap}
	\caption{ Heatmap of the cosine similarity of the vectors
          representing some of the tokens used in the word frequency
          experiment.  The words other than \word{the} and \word{VOID}
          were chosen randomly.  }
	\label{word-frequency-experiment-heatmap}
\end{figure}

\begin{table}
	\input{cosine-similarity.tex}
	\caption{ Words in the original vocabulary most similar to the
          tokens \word{THE\_i}, and their cosine similarity with the
          most frequent such token, \word{THE\_1}.  It is apparent from
          the nearest neighbour list that the vectors of the
          low-frequency tokens have not been adequately trained.  }
	\label{word-frequency-experiment-cosine-similarity}
\end{table}

\begin{sidewaysfigure*}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{ Vector length vs.\ frequency for tokens derived from a
          few words chosen at random.  For each word, tokens of varying
          frequency but with the co-occurrence distribution of that word
          were inserted into the corpus, as described in
          Section~\ref{WFVE}.  The vectors are obtained from the first
          synaptic layer, syn0, of the word2vec neural network.  The
          vectors obtained from the second layer, syn1neg, are included
          for completeness.  Legend entries are ordered by vector length
          of the left-most data point in the syn0 plot, descending.}
	\label{fig:word-frequency-experiment-graph}
\end{sidewaysfigure*}
\end{section}

@@

For each of these words, the replacement procedure of
Section~\ref{replacement-procedure} is performed using the
distribution~(\ref{distro2}) with $n=7$.  For every replacement token
(e.g. \word{CAT\_i}), additional occurrences of this token are
interspersed uniformly at random throughout the corpus, such that the
final frequency of the replacement token is $2/n$ times that of the
original word \word{cat}.  For example, if the original word \word{cat}
occurred $1000$ times, then after the replacement procedure,
\word{CAT\_2} occurs approximately $238$ times, so a further
(approximately) $2/7 \times 1000 - 238 \approx 48$ random occurrences of
\word{CAT\_2} are interspersed throughout the corpus.  In this way, the
word \word{cat} is removed from the corpus and replaced with a family of
tokens \word{CAT\_i}, $1 \leqslant i \leqslant 7$.  These tokens all
have the same frequency, but their co-occurrence distributions, while
based on that of \word{cat}, have an increasing amount of noise.
Specifically, the proportion of noise for the $i$th token is
%
$$ 1 - \frac{n}{2} P_n(i) = \frac{i-1}{n-1} , \;\; \textrm{ or } \;\; 0,
\frac{1}{n-1}, \frac{2}{n-1}, \ldots, 1 \;\; \textrm{ for } \;\; i =
1,2, \ldots, n , 
$$ 
%
which is evenly distributed.  The first token contains no noise at all,
while the last token stands for pure noise.  The particular choice of
$n$ assures a reasonable coverage of the interval $[0,1]$.  Other
parameter values (or indeed other distributions) could, of course, have
been used equally well.

Figure~\ref{fig:co-occurrence-noise-experiment-text} illustrates the
effect of this modification in the case where the only word chosen is
\word{cat}.  The original text in this case concerned both cats and
dogs.  Notice that the word \word{cat} has been replaced entirely in the
cats section by \word{CAT\_i} and, moreover, that these same tokens
appear also in the dogs section. These occurrences (and additionally,
with probability, some occurrences from the cats section) constitute
noise.

@@

\begin{figure}
	\begin{mdframed}
	\input{cooccurrence-noise-experiment.tex}
	\end{mdframed}
	\caption{Example sentences modified for the co-occurrence noise
          experiment, where the word \word{cat} was chosen for
          replacement.  The tokens were generated using the distribution
          (\ref{distro2}) with $n=7$.}
	\label{fig:co-occurrence-noise-experiment-text}
\end{figure}

>>>>

In this paper, we wish to study in detail to what extent these two
factors determine word vectors.  For a given corpus, both term
frequency and co-occurrence are, of course, fixed and it is not obvious
how to unravel these dependencies in an unambiguous, objective manner.
In particular, it is difficult to establish the distinctiveness of the
contexts in which a word is used.  To overcome these problems, we
propose to modify the training corpus in a controlled fashion.  To
this end, we insert new pseudowords into the corpus with varying frequencies
and varying levels of noise in their co-occurrence distributions.  By
modeling the frequency and co-occurrence distributions of these pseudowords
on existing words in the corpus, we are able to study their effect on
word vectors independently of one another.  We can thus study a family
of pseudowords that all appear in the same context, but with different
frequencies, or study a family of pseudowords that all have the same
frequency, but appear in a different number of contexts.  Starting from
the limited number of contexts in which a word appears in the original
corpus, we can increase this number by interspersing the word in arbitrary
contexts at random.  The word thus looses its significance in a
controlled way.  Although we present our approach using the word2vec
CBOW model, these and related experiments could equally well be carried
out for other word embedding methods such as the word2vec skip-gram
model \cite{DistRepns,EfficientEstimation}, GloVe
\cite{pennington2014glove}, and SENNA \cite{collobert-2011}.

We show that the length of the word vectors generated by the CBOW model
depends more or less linearly on both word frequency and level of noise
in the co-occurrence distribution of the word.  In both cases, the
coefficient of linearity depends upon the word.  If the co-occurrence
distribution is fixed, then word vector length increases with word
frequency.  If, on the other hand, word frequency is held constant, then
word vector length decreases as the level of noise in the co-occurrence
distribution of the word is increased.  We show furthermore that the
direction of the word vectors is independent of word frequency and only
moderately depends upon co-occurrence noise.  When noise is added to the
co-occurrence distribution of a word, the corresponding vector smoothly
interpolates between the original word vector and a small vector
perpendicular to it that represents a word with pure noise in its
co-occurrence distribution.  Surprisingly, this special vector in the feature
space, obtained by interspersing a pseudoword uniformly at random throughout
the corpus with a frequency sufficiently large to sample all contexts, is
non-zero.

@@

\begin{section}{Related work}\label{related-work}
In recent work on the evaluation of word embeddings, Schnabel, Labutov, Mimno
and Joachims \cite{schnabelemnlp2015} trained logistic regression models to
predict whether a word was rare or frequent given only the direction of its
word vector.  The prediction accuracy was then measured as a function of word
embedding method and of the threshold for word rarity.  It was found in the
case of word2vec CBOW that word vector direction could be used to distinguish
very rare words from all other words. The authors claimed further that word
vector direction must encode frequency directly, and not indirectly via
semantic information.  Our experimental finding that word vector direction is
independent of word frequency contradicts this claim in the case of word2vec
CBOW.

@@

\begin{section}{Corpus and model}\label{corpus-and-model} Our training data is
	built from the Wikipedia data dump from October 2013.  To remove the
	bulk of robot-generated pages from the training data, only pages with
	at least 20 monthly page views are retained.\footnote{For further
	justification and to obtain the dataset, see\\ \blogpost} Stubs and
	disambiguation pages are also removed, leaving 463 thousand pages with
	a total of 482 million words.  Punctuation marks and numbers were
	removed from the pages and all words were lower-cased.  Word
	frequencies are summarised in Table~\ref{word-occurrences-table}.  This
	base corpus is then modified as described in Sections~\ref{WFVE} and
	\ref{CNVE}.  For recognisability, the pseudowords inserted into the
	corpus are upper-cased.

@@

\begin{subsection}{Replacement procedure}\label{replacement-procedure}
In the experiments detailed below, we modify the corpus in a controlled
manner by introducing pseudowords into the corpus via a replacement
procedure.  For the frequency experiment, the procedure is as follows.
Consider a word, say \word{cat}.  For each occurrence of this word, a
sample $i$, $1 \leqslant i \leqslant n$ is drawn from a truncated
geometric distribution, and that occurrence of the word \word{cat} is
replaced with the pseudoword \word{CAT\_i}.  In this way, the word \word{cat}
is replaced throughout the corpus by a family of pseudowords with varying
frequencies but approximately the same co-occurrence distribution as
\word{cat}.  That is, all these pseudowords are used in roughly the same
contexts as the original word.

The geometric distribution is truncated to limit the number of pseudowords
inserted into the corpus.  For any choice $0 < p < 1$ and maximum value
$n > 0$, the truncated geometric distribution is given by the
probability density function
\begin{equation} 
\label{distro}
P_{p, n} (i) = \frac{p^{i-1}  (1-p)}{1 - p^n}, \qquad 1
\leqslant i \leqslant n.
\end{equation} 
The factor in the denominator, which tends to unity in the limit $n \to
\infty$, assures proper normalisation.  We have chosen this distribution
because the probabilities decay exponentially base $p$ as a function of
$i$.  Of course, other distributions might equally well have been chosen
for the experiments.

@@

\begin{section}{Varying word frequency}\label{WFVE}
In this first experiment, we investigate the effect of word frequency on
the word embedding.  Using the replacement procedure, we introduce a
small number of families of pseudowords into the corpus.  The pseudowords in each
family vary in frequency but, replacing a single word, all share a
common co-occurrence distribution.  This allows us to study the role of
word frequency in isolation, everything else being kept equal.  We
consider two types of pseudowords.

\begin{subsection}{Pseudowords derived from existing words}\label{WFVEexisting}
We choose uniformly at random a small number of words from the
unmodified vocabulary for our experiment.  In order that the inserted
pseudowords do not have too low a frequency, only words which occur at least
10 thousand times are chosen.  We also include the high-frequency
stopword \word{the} for comparison.  Table~\ref{word-frequency-counts}
lists the words chosen for this experiment along with their frequencies.

The replacement procedure of Section~\ref{replacement-procedure} is then
performed for each of these words, using a geometric decay rate of $p =
\tfrac{1}{2}$, and maximum value $n=20$, so that the 1st pseudoword is
inserted with a probability of about $0.5$, the 2nd with a probability of
about $0.25$, and so on.  This value of $p$ is one of a range of values
that ensure that, for each word, multiple pseudowords will be inserted with a
frequency sufficient to survive the low-frequency cut-off of
128.  A maximum value $n=20$ suffices for this choice of $p$, since
$2^{20 + \log_2{128}}$ exceeds the maximum frequency of any word in
the corpus.  Figure~\ref{fig:word-frequency-experiment-text-cat}
illustrates the effect of these modifications on a sample text, with a
family of pseudowords \word{CAT\_i}, derived from the word \word{cat}.
Notice that all occurrences of the word \word{cat} have been replaced
with the pseudowords \word{CAT\_i}.

@@

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{Example sentences modified in the word frequency
          experiment as per Section~\ref{WFVEexisting}, where the word
          \word{cat} is replaced with pseudowords \word{CAT\_i} using the
          truncated geometric distribution (\ref{distro}) with
          $p=\tfrac{1}{2}$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}
\label{WFVEmeaningless}
Whereas the pseudowords introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we intersperse the word uniformly at random throughout the corpus
so that its relative frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the unconditional word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure~\ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher relative frequency
of $0.05$ is used instead for illustrative purposes.
%Although the probability of inserting the pseudoword
%\word{VOID\_1} is twice that of inserting the pseudoword \word{VOID\_2} and four
%times that of inserting the pseudoword \word{VOID\_3}, these three pseudowords all
%appear once in this small piece of text.

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{The same example sentences as in
          Figure~\ref{fig:word-frequency-experiment-text-cat} where
          instead of the word \word{cat} now the meaningless word
          \word{VOID} is replaced with pseudowords \word{VOID\_i}.  For
          illustrative purposes, the meaningless word \word{VOID} was
          here interspersed with a relative frequency of
          $0.05$. }
	\label{fig:word-frequency-experiment-text-void}
\end{figure}
\end{subsection}

@@

\subsubsection{Word frequency and vector direction}\label{WFVE-direction}
Figure~\ref{word-frequency-experiment-heatmap} shows the cosine
similarity of pairs of vectors representing some of the pseudowords used in
this experiment.  Recall that the cosine similarity measures the extent
to which two vectors have the same direction, taking a maximum value of
$1$ and a minimum value of $-1$.  The number of different pseudowords
associated with an experiment word is the number of times that its
frequency can be halved and remain above the low-frequency cut-off of
$128$.

Consider first the words other than \word{the} and \word{VOID}.  These
words typically occur in a limited number of distinct contexts, and so a
limited number of samples is required to capture their co-occurrence
distributions.  We therefore expect that the vectors representing the
pseudowords derived from any of these words can be properly learned from the
modified corpus, even though their combined frequency equals the
frequency of that word in the original corpus.  The red blocks in
Figure~\ref{word-frequency-experiment-heatmap} clearly demonstrate that
the direction of these vectors is constant for each family of pseudowords.
In other words, the direction of these vectors is only determined by the
original word and is independent of how often that word appears,
provided sufficient samples have been processed for the word vector to
settle in its equilibrium direction.

The artificial word \word{VOID}, on the other hand, occurs by
construction in every context.  A much higher number of samples is
therefore required to capture its co-occurrence distribution, and
thereby to learn its word vector.  The same is true, but to a lesser
extent, for the stopword \word{the}.  The difficulty of learning the
vector representing \word{VOID\_i} or \word{THE\_i} increases with $i$,
for the pseudowords occur increasingly less often.  This is consistent with
Figure~\ref{word-frequency-experiment-heatmap}, where it is apparent
that the vectors representing \word{VOID\_i} for small $i$, say $i = 1,
\dots, 5$, roughly share a common direction, while this is no longer
true for higher values of $i$.  The same trend is apparent, although
less extremely so, for the word \word{the}.  The vectors for the pseudowords
\word{THE\_i} share a common direction for values of $i$ as high as
$14$, before they start changing direction.  We conclude that these
deviations are artifacts of poor learning due to an insufficient number
of occurrences of these pseudowords.
Table~\ref{word-frequency-experiment-cosine-similarity}, showing the
most similar words to each pseudoword \word{THE\_i}, supports this further.
In the discussions below, the pseudowords \word{VOID\_i} and \word{THE\_i}
whose vectors have a cosine similarity less than $0.8$ with
\word{VOID\_1} and \word{THE\_1}, respectively, are therefore discarded.

@@

\subsubsection{Word frequency and vector length}
We next consider the effect of frequency on word vector length.
Throughout, we measure vector length using the Euclidean norm.
Figure~\ref{fig:word-frequency-experiment-graph} shows this relation for
individual words, both for the word vectors, represented by the weights
of the first synaptic layer, syn0, in the word2vec neural network, and
for the vectors represented by the weights of the second synaptic layer,
syn1neg.  We include the latter, which are typically ignored, for
completeness.  Each line corresponds to a single word, and the points on
each line indicate the frequency and vector length of the pseudowords derived
from that word.  For example, the six points on the line corresponding
to the word \word{protestant} are labeled, from right to left, by the
pseudowords \word{PROTESTANT\_1}, \word{PROTESTANT\_2}, \dots,
\word{PROTESTANT\_6}.  Again, the number of points on the line is
determined by the frequency of the original word.  For example, the
frequency of the word \word{protestant} can be halved at most $6$ times
so that the frequency of the last pseudoword is still above the low-frequency
cut-off.  Because all the points on a line share the same co-occurrence
distribution, the left panel in
Figure~\ref{fig:word-frequency-experiment-graph} demonstrates
conclusively that length does indeed depend on frequency directly.
Moreover, this relation is seen to be approximately linear for each word
considered.  Notice also that the relative positions of the lengths of
the word vectors associated with the experiment words are roughly
independent of the frequency band, i.e., the plotted lines rarely cross.

Notice the inflection of the syn1neg curve in
Figure~\ref{fig:word-frequency-experiment-graph} for \word{the} at
approximately $10^6$.  We have no explanation of this.  Also observe
that the lengths of the vectors representing the meaningless pseudowords
\word{VOID\_i} are approximately constant (about $2.5$).  Since we
already found the direction to be also constant, it is sensible to speak
of the word vector of \word{VOID} irrespective of its frequency.  In
particular, the vector of the pseudoword \word{VOID\_1} may be taken as an approximation.

\begin{figure}
	\includegraphics[scale=0.5]{word-frequency-experiment-heatmap}
	\caption{ Heatmap of the cosine similarity of the vectors
          representing some of the pseudowords used in the word frequency
          experiment.  The words other than \word{the} and \word{VOID}
          were chosen randomly.  }
	\label{word-frequency-experiment-heatmap}
\end{figure}

\begin{table}
	\input{cosine-similarity.tex}
	\caption{ Words in the original vocabulary most similar to the
          pseudowords \word{THE\_i}, and their cosine similarity with the
          most frequent such pseudoword, \word{THE\_1}.  It is apparent from
          the nearest neighbour list that the vectors of the
          low-frequency pseudowords have not been adequately trained.  }
	\label{word-frequency-experiment-cosine-similarity}
\end{table}

\begin{sidewaysfigure*}
	\centering{\includegraphics[scale=0.6]{word-frequency-experiment-graph}}
	\caption{ Vector length vs.\ frequency for pseudowords derived from a
          few words chosen at random.  For each word, pseudowords of varying
          frequency but with the co-occurrence distribution of that word
          were inserted into the corpus, as described in
          Section~\ref{WFVE}.  The vectors are obtained from the first
          synaptic layer, syn0, of the word2vec neural network.  The
          vectors obtained from the second layer, syn1neg, are included
          for completeness.  Legend entries are ordered by vector length
          of the left-most data point in the syn0 plot, descending.}
	\label{fig:word-frequency-experiment-graph}
\end{sidewaysfigure*}
\end{section}

@@

For each of these words, the replacement procedure of
Section~\ref{replacement-procedure} is performed using the
distribution~(\ref{distro2}) with $n=7$.  For every replacement pseudoword
(e.g. \word{CAT\_i}), additional occurrences of this pseudoword are
interspersed uniformly at random throughout the corpus, such that the
final frequency of the replacement pseudoword is $2/n$ times that of the
original word \word{cat}.  For example, if the original word \word{cat}
occurred $1000$ times, then after the replacement procedure,
\word{CAT\_2} occurs approximately $238$ times, so a further
(approximately) $2/7 \times 1000 - 238 \approx 48$ random occurrences of
\word{CAT\_2} are interspersed throughout the corpus.  In this way, the
word \word{cat} is removed from the corpus and replaced with a family of
pseudowords \word{CAT\_i}, $1 \leqslant i \leqslant 7$.  These pseudowords all
have the same frequency, but their co-occurrence distributions, while
based on that of \word{cat}, have an increasing amount of noise.
Specifically, the proportion of noise for the $i$th pseudoword is
%
$$ 1 - \frac{n}{2} P_n(i) = \frac{i-1}{n-1} , \;\; \textrm{ or } \;\; 0,
\frac{1}{n-1}, \frac{2}{n-1}, \ldots, 1 \;\; \textrm{ for } \;\; i =
1,2, \ldots, n , 
$$ 
%
which is evenly distributed.  The first pseudoword contains no noise at all,
while the last pseudoword stands for pure noise.  The particular choice of
$n$ assures a reasonable coverage of the interval $[0,1]$.  Other
parameter values (or indeed other distributions) could, of course, have
been used equally well.

Figure~\ref{fig:co-occurrence-noise-experiment-text} illustrates the
effect of this modification in the case where the only word chosen is
\word{cat}.  The original text in this case concerned both cats and
dogs.  Notice that the word \word{cat} has been replaced entirely in the
cats section by \word{CAT\_i} and, moreover, that these same pseudowords
appear also in the dogs section. These occurrences (and additionally,
with probability, some occurrences from the cats section) constitute
noise.

@@

\begin{figure}
	\begin{mdframed}
	\input{cooccurrence-noise-experiment.tex}
	\end{mdframed}
	\caption{Example sentences modified for the co-occurrence noise
          experiment, where the word \word{cat} was chosen for
          replacement.  The pseudowords were generated using the distribution
          (\ref{distro2}) with $n=7$.}
	\label{fig:co-occurrence-noise-experiment-text}
\end{figure}

\begin{subsection}{Experimental results}
%
Figure~\ref{fig:co-occurrence-noise-heatmap} shows the cosine similarity
of pairs of vectors representing some of the pseudowords used in this
experiment.  Remember that the first pseudoword ($i=1$) in a family is
without noise in its co-occurrence distribution, while the last one
($i=n$, with $n=7$) stands for pure noise and has therefore no relation
anymore with the word it derives from.  The figure demonstrates that the
vectors within a family only moderately deviate from the original
direction defined by the first pseudoword ($i=1$) when noise is added to the
co-occurrence distribution.  For $1<i<7$, the deviation typically
increases with the proportion of noise.  The vector of the last pseudoword
($i=n$), associated with pure noise, is seen within each of the families
to point in a completely different direction, more or less perpendicular
to the original one.  To understand this interpolating behavior, recall
from Section~\ref{WFVE-results} that the vector for the entirely
meaningless word \word{VOID} is small but non-zero.  Since the noise
distribution coincides with the co-occurrence distribution of
\word{VOID}, the vectors for the experiment words must tend to the word
vector for \word{VOID} as the proportion of noise in their co-occurrence
distributions approaches $1$.
This convergence to a common point is only indistinctly apparent in
Figure~\ref{fig:co-occurrence-noise-heatmap} as the frequency of the experiment
pseudowords is insufficient to sample the full variety of the contexts of
\word{VOID}, i.e.  all contexts (see Section \ref{WFVE-direction}).
